A Step-Indexed Logic in Agda

### 2023-06-21 papers/cpp2024/paper.tex
check

%%
%% This is file `sample-acmsmall.tex',
%% generated with the docstrip utility.
%%
%% The original source files were:
%%
%% samples.dtx  (with options: `acmsmall')
%% 
%% IMPORTANT NOTICE:
%% 
%% For the copyright see the source file.
%% 
%% Any modified versions of this file must be renamed
%% with new filenames distinct from sample-acmsmall.tex.
%% 
%% For distribution of the original source see the terms
%% for copying and modification in the file samples.dtx.
%% 
%% This generated file may be distributed as long as the
%% original source files, as listed above, are part of the
%% same distribution. (The sources need not necessarily be
%% in the same archive or directory.)
%%
%%
%% Commands for TeXCount
%TC:macro \cite [option:text,text]
%TC:macro \citep [option:text,text]
%TC:macro \citet [option:text,text]
%TC:envir table 0 1
%TC:envir table* 0 1
%TC:envir tabular [ignore] word
%TC:envir displaymath 0 word
%TC:envir math 0 word
%TC:envir comment 0 0
%%
%%
%% The first command in your LaTeX source must be the \documentclass
%% command.
%%
%% For submission and review of your manuscript please change the
%% command to \documentclass[manuscript, screen, review]{acmart}.
%%
%% When submitting camera ready or to TAPS, please change the command
%% to \documentclass[sigconf]{acmart} or whichever template is required
%% for your publication.
%%
%%
\documentclass[acmsmall]{acmart}
\usepackage{semantic}

@@

@@

%%
%% The abstract is a short summary of the work to be presented in the
%% article.
\begin{abstract}
  Step-indexed logical relations are an important tool in the
  metatheory of programming
  languages~\citep{Appel:2001aa,Ahmed:2006aa}, and their application
  can be streamlined by use of modal logics such as
  LSLR~\citep{Dreyer:2011wl}.  The Iris system~\citep{JUNG:2018aa}
  provides support for reasoning about step-indexed logical relations
  in the Coq proof assistant, but there is not yet analogous support
  for the Agda proof assistant. This paper describes an embedding of
  step-indexed logic in Agda and an application of the logic to a
  semantic type safety proof for an internal language for the
  Gradually-Typed Lambda Calculus.
\end{abstract}

@@

The author is interested in step-indexed logical relations for
purposes of developing mechanized proofs in Agda of the Gradual
Guarantee~\citep{Siek:2015ac} and Parametricity~\citep{REYNOLDS74C}
for gradually-typed polymorphic
languages~\citep{Ahmed:2011fk,Ahmed:2017aa,Igarashi:2017aa,New:2019ab,Labrada:2020tk}.
To streamline such developments in Agda, we need support for a
step-indexed logic in Agda. This paper fills that need by showing how
to build a shallow embedding of a step-indexed logic in Agda, a system
that we abbreviate as SIL. To evaluate the utility of SIL, we present
a case study of proving semantic type safety of a simple cast
calculus, i.e., an internal language for the Gradually-Typed Lambda
Calculus (GTLC)~\citep{Siek:2006bh,Siek:2007qy}. The outcome of that
study is that SIL streamlines the definition of logical relations and
is helpful in many of the proofs, but there is room for improvement
regarding the ergonomics of interactive proof development.
Limitations in Agda's inference algorithm sometimes leads to verbosity
in SIL proofs. It is worth noting that the authors have also proved
the Gradual Guarantee for the GTLC using SIL and the lesssons from
that experience confirm the findings in this paper.

TODO: ROAD MAP

\section{A Step-Indexed Logic (SIL)}

Figure~\ref{fig:SIL}

\begin{figure}
  \raggedright
Syntax
  \begin{syntax}
    \text{Propositions} & P,Q &::= & \top \mid \bot \mid P \land Q
    \mid P \lor Q \mid P \Rightarrow Q \mid \forall X.P \mid \exists X.P
    \mid \overline{M} \in R\\
    \text{Relations} & R,S & ::= & r \mid \mu r. R \mid \{ \overline{x} \mid P \}
  \end{syntax}
Meaning of Propositions
\begin{align*}
  \sem{ P } \rho\; 0 &= \top \\
  \sem{ \top } \rho(suc \app n) &= \top \\
  \sem{ \bot } \rho (suc \app n) &= \bot \\
  \sem{ P \land Q } \rho (suc \app n) &=
    \sem{ P } \rho n \land \sem{ Q } \rho n \\
  \sem{ P \lor Q } \rho (suc \app n) &=
    \sem{ P } \rho n \lor \sem{ Q } \rho n \\
  \sem{ P \Rightarrow Q } \rho (suc \app n) &=
    \forall k \leq suc \app n,
    \sem{ P } \rho k \implies \sem{ Q } \rho k \\
  \sem{ \forall X.P} \rho \app (suc \app n) &= \forall a, \sem{P[a/X]} \rho (suc \app n)\\
  \sem{ \exists X.P} \rho \app (suc \app n) &= \exists a, \sem{P[a/X]} \rho (suc \app n)\\
  \sem{ \overline{M} \in R } \rho (suc \app n) &=
    \sem{ R } \rho (suc \app n) \app \overline{M}\\
  \sem{ \rhd P } \rho (suc \app n) &=
    \sem{ P } \rho n 
\end{align*}
Meaning of Relations
\begin{align*}
  \sem{ R } \rho\, 0 \app \overline{M} &= \top \\
  \sem{ r } \rho \app (suc \app n) \app \overline{M} &=
    \rho(r) (suc \app n) \app \overline{M} \\
  \sem{ \mu r. R } \rho \app (suc \app n)\app \overline{M} &=
  F^{suc \app n}(\top) (suc \app n) \app \overline{M} \\
  & \text{ where } F(X) = \sem{ R } \rho(r := X) \\
  \sem{ \{ \overline{x} \mid P \} } \rho (suc \app n) \app \overline{M} &= 
    \sem{ P[\overline{x} := \overline{M}] } \rho (suc \app n)
\end{align*}
  \caption{Step-Indexed Logic (SIL)}
  \label{fig:SIL}
\end{figure}

\section{Case Study: Semantic Type Safety of a Cast Calculus}

Figure~\ref{fig:CC}

\begin{figure}
  \raggedright Syntax
  \begin{syntax}
    \text{Types} & A,B,C,D &::=& \Bool \mid A \to B \mid \dyn \\
    \text{Ground Types} & G,H & ::= & \mathtt{bool} \mid \dyn \to \dyn \\
    \text{Booleans} & b & \in & \mathbb{B} \\
    \text{Terms} & L,M,N & ::= & x \mid b \mid \lam{x}N \mid L \app M
      \mid \inj{M}{G} \mid \proj{M}{H} \mid \mathtt{error} \\
      \text{Values} & V,W & ::= & \lam{x}N \mid \inj{V}{G}
  \end{syntax}
  Type System
  \begin{gather*}
    \inference{}{\Gamma \vdash x : \Gamma(x)}
    \quad
    \inference{}{\Gamma \vdash b : \Bool}
    \quad
    \inference{\Gamma, x : A \vdash N : B}
              {\Gamma \vdash \lam{x} N : A \to B}
    \quad              
    \inference{\Gamma \vdash L : A \to B & \Gamma \vdash M : A}
          {\Gamma \vdash L \app M : B} \\[2ex]
   \inference{\Gamma \vdash M : G}
          {\Gamma \vdash \inj{M}{G} : \dyn}
   \quad
   \inference{\Gamma \vdash M : \dyn}
             {\Gamma \vdash \proj{M}{H} : H}
   \quad
   \inference{}
             {\Gamma \vdash \mathtt{error} : A}
  \end{gather*}
Reduction
\begin{align*}
  (\lam{x} N)\app W & \reduce N[x:=W] \\
  \proj{\inj{V}{G}}{G} & \reduce V \\
  \proj{\inj{V}{G}}{H} & \reduce \texttt{error}
       \qquad\qquad\qquad\text{if } G \neq H 
\end{align*}

  \caption{Internal language (cast calculus) for the Gradually-Typed
    Lambda Calculus.}
  \label{fig:CC}
\end{figure}

\begin{align*}
    \semE{\_} &: \mathsf{Type} \to \mathsf{Term} \to \mathsf{Set} \\
    \semE{A} M &= (\semV{A} M \lor \mathsf{reducible}\ M \lor M=\mathtt{blame})
                \land (\forall N, (M \reduce N) \Rightarrow \rhd \semE{A} N)
\end{align*}

>>>>

%%
%% This is file `sample-acmsmall.tex',
%% generated with the docstrip utility.
%%
%% The original source files were:
%%
%% samples.dtx  (with options: `acmsmall')
%% 
%% IMPORTANT NOTICE:
%% 
%% For the copyright see the source file.
%% 
%% Any modified versions of this file must be renamed
%% with new filenames distinct from sample-acmsmall.tex.
%% 
%% For distribution of the original source see the terms
%% for copying and modification in the file samples.dtx.
%% 
%% This generated file may be distributed as long as the
%% original source files, as listed above, are part of the
%% same distribution. (The sources need not necessarily be
%% in the same archive or directory.)
%%
%%
%% Commands for TeXCount
%TC:macro \cite [option:text,text]
%TC:macro \citep [option:text,text]
%TC:macro \citet [option:text,text]
%TC:envir table 0 1
%TC:envir table* 0 1
%TC:envir tabular [ignore] word
%TC:envir displaymath 0 word
%TC:envir math 0 word
%TC:envir comment 0 0
%%
%%
%% The first command in your LaTeX source must be the \documentclass
%% command.
%%
%% For submission and review of your manuscript please change the
%% command to \documentclass[manuscript, screen, review]{acmart}.
%%
%% When submitting camera ready or to TAPS, please change the command
%% to \documentclass[sigconf]{acmart} or whichever template is required
%% for your publication.
%%
%%
\documentclass[acmsmall]{acmart}
\usepackage{semantic}
\usepackage{agda}
\usepackage{MnSymbol}

@@

\usepackage{newunicodechar}
\newunicodechar{λ}{\ensuremath{\mathnormal\lambda}}
\newunicodechar{←}{\ensuremath{\mathnormal\from}}
\newunicodechar{→}{\ensuremath{\mathnormal\to}}
\newunicodechar{∀}{\ensuremath{\mathnormal\forall}}
\newunicodechar{∃}{\ensuremath{\exists}}
\newunicodechar{⊥}{\ensuremath{\mathnormal\bot}}
\newunicodechar{⊤}{\ensuremath{\mathnormal\top}}
\newunicodechar{∷}{\ensuremath{\mathnormal::}}
\newunicodechar{ℕ}{\ensuremath{\mathbb N}}
\newunicodechar{≤}{\ensuremath{\leq}}
\newunicodechar{∸}{\ensuremath{\div}}
\newunicodechar{′}{\ensuremath{'}}
\newunicodechar{⇒}{\ensuremath{\Rightarrow}}
\newunicodechar{₁}{\ensuremath{_1}}
\newunicodechar{₂}{\ensuremath{_2}}
\newunicodechar{Σ}{\ensuremath{\Sigma}}
\newunicodechar{⊎}{\ensuremath{\cupdot}}
\newunicodechar{ᵖ}{\ensuremath{^p}}
\newunicodechar{≡}{\ensuremath{\equiv}}
\newunicodechar{≢}{\ensuremath{\not\equiv}}
\newunicodechar{∘}{\ensuremath{\circ}}
\newunicodechar{ᵒ}{\ensuremath{^\circ}}
\newunicodechar{▷}{\ensuremath{\rhd}}

@@

%%
%% The abstract is a short summary of the work to be presented in the
%% article.
\begin{abstract}
  Step-indexed logical relations are an important tool in the
  metatheory of programming
  languages~\citep{Appel:2001aa,Ahmed:2006aa}, and their application
  can be streamlined by use of modal logics such as
  LSLR~\citep{Dreyer:2011wl}.  The Iris system~\citep{JUNG:2018aa}
  provides support for reasoning about step-indexed logical relations
  in the Coq proof assistant, but there is not yet analogous support
  for the Agda proof assistant. This paper describes an embedding of
  a step-indexed logic in Agda and an application of the logic to a
  semantic type safety proof for an internal language for the
  Gradually-Typed Lambda Calculus.
\end{abstract}

@@

The author is interested in step-indexed logical relations for
purposes of developing mechanized proofs in Agda of the Gradual
Guarantee~\citep{Siek:2015ac} and Parametricity~\citep{REYNOLDS74C}
for gradually-typed polymorphic
languages~\citep{Ahmed:2011fk,Ahmed:2017aa,Igarashi:2017aa,New:2019ab,Labrada:2020tk}.
To streamline such developments in Agda, we need support for a
step-indexed logic in Agda. This paper fills that need by showing how
to build a shallow embedding of a step-indexed logic in Agda, a system
that we abbreviate as SIL. To evaluate the utility of SIL, we present
a case study of proving semantic type safety of a simple cast
calculus, i.e., an internal language for the Gradually-Typed Lambda
Calculus (GTLC)~\citep{Siek:2006bh,Siek:2007qy}. We find that SIL
streamlines the definition of logical relations and is helpful in many
of the proofs, but there is room for improvement regarding the
ergonomics of interactive proof development using SIL.  Limitations in
Agda's inference algorithm sometimes leads to verbosity in SIL
proofs. It is worth noting that the authors have also proved the
Gradual Guarantee for the GTLC using SIL and the lesssons from that
experience confirm the findings in this paper.

%TODO: ROAD MAP

%\section{A Step-Indexed Logic (SIL)}

%% Figure~\ref{fig:SIL}

%% \begin{figure}
%%   \raggedright
%% Syntax
%%   \begin{syntax}
%%     \text{Propositions} & P,Q &::= & \top \mid \bot \mid P \land Q
%%     \mid P \lor Q \mid P \Rightarrow Q \mid \forall X.P \mid \exists X.P
%%     \mid \overline{M} \in R\\
%%     \text{Relations} & R,S & ::= & r \mid \mu r. R \mid \{ \overline{x} \mid P \}
%%   \end{syntax}
%% Meaning of Propositions
%% \begin{align*}
%%   \sem{ P } \rho\; 0 &= \top \\
%%   \sem{ \top } \rho(suc \app n) &= \top \\
%%   \sem{ \bot } \rho (suc \app n) &= \bot \\
%%   \sem{ P \land Q } \rho (suc \app n) &=
%%     \sem{ P } \rho n \land \sem{ Q } \rho n \\
%%   \sem{ P \lor Q } \rho (suc \app n) &=
%%     \sem{ P } \rho n \lor \sem{ Q } \rho n \\
%%   \sem{ P \Rightarrow Q } \rho (suc \app n) &=
%%     \forall k \leq suc \app n,
%%     \sem{ P } \rho k \implies \sem{ Q } \rho k \\
%%   \sem{ \forall X.P} \rho \app (suc \app n) &= \forall a, \sem{P[a/X]} \rho (suc \app n)\\
%%   \sem{ \exists X.P} \rho \app (suc \app n) &= \exists a, \sem{P[a/X]} \rho (suc \app n)\\
%%   \sem{ \overline{M} \in R } \rho (suc \app n) &=
%%     \sem{ R } \rho (suc \app n) \app \overline{M}\\
%%   \sem{ \rhd P } \rho (suc \app n) &=
%%     \sem{ P } \rho n 
%% \end{align*}
%% Meaning of Relations
%% \begin{align*}
%%   \sem{ R } \rho\, 0 \app \overline{M} &= \top \\
%%   \sem{ r } \rho \app (suc \app n) \app \overline{M} &=
%%     \rho(r) (suc \app n) \app \overline{M} \\
%%   \sem{ \mu r. R } \rho \app (suc \app n)\app \overline{M} &=
%%   F^{suc \app n}(\top) (suc \app n) \app \overline{M} \\
%%   & \text{ where } F(X) = \sem{ R } \rho(r := X) \\
%%   \sem{ \{ \overline{x} \mid P \} } \rho (suc \app n) \app \overline{M} &= 
%%     \sem{ P[\overline{x} := \overline{M}] } \rho (suc \app n)
%% \end{align*}
%%   \caption{Step-Indexed Logic (SIL)}
%%   \label{fig:SIL}
%% \end{figure}

%\section{Case Study: Semantic Type Safety of a Cast Calculus}

%% Figure~\ref{fig:CC}

%% \begin{figure}
%%   \raggedright Syntax
%%   \begin{syntax}
%%     \text{Types} & A,B,C,D &::=& \Bool \mid A \to B \mid \dyn \\
%%     \text{Ground Types} & G,H & ::= & \mathtt{bool} \mid \dyn \to \dyn \\
%%     \text{Booleans} & b & \in & \mathbb{B} \\
%%     \text{Terms} & L,M,N & ::= & x \mid b \mid \lam{x}N \mid L \app M
%%       \mid \inj{M}{G} \mid \proj{M}{H} \mid \mathtt{error} \\
%%       \text{Values} & V,W & ::= & \lam{x}N \mid \inj{V}{G}
%%   \end{syntax}
%%   Type System
%%   \begin{gather*}
%%     \inference{}{\Gamma \vdash x : \Gamma(x)}
%%     \quad
%%     \inference{}{\Gamma \vdash b : \Bool}
%%     \quad
%%     \inference{\Gamma, x : A \vdash N : B}
%%               {\Gamma \vdash \lam{x} N : A \to B}
%%     \quad              
%%     \inference{\Gamma \vdash L : A \to B & \Gamma \vdash M : A}
%%           {\Gamma \vdash L \app M : B} \\[2ex]
%%    \inference{\Gamma \vdash M : G}
%%           {\Gamma \vdash \inj{M}{G} : \dyn}
%%    \quad
%%    \inference{\Gamma \vdash M : \dyn}
%%              {\Gamma \vdash \proj{M}{H} : H}
%%    \quad
%%    \inference{}
%%              {\Gamma \vdash \mathtt{error} : A}
%%   \end{gather*}
%% Reduction
%% \begin{align*}
%%   (\lam{x} N)\app W & \reduce N[x:=W] \\
%%   \proj{\inj{V}{G}}{G} & \reduce V \\
%%   \proj{\inj{V}{G}}{H} & \reduce \texttt{error}
%%        \qquad\qquad\qquad\text{if } G \neq H 
%% \end{align*}

%%   \caption{Internal language (cast calculus) for the Gradually-Typed
%%     Lambda Calculus.}
%%   \label{fig:CC}
%% \end{figure}

%% \begin{align*}
%%     \semE{\_} &: \mathsf{Type} \to \mathsf{Term} \to \mathsf{Set} \\
%%     \semE{A} M &= (\semV{A} M \lor \mathsf{reducible}\ M \lor M=\mathtt{blame})
%%                 \land (\forall N, (M \reduce N) \Rightarrow \rhd \semE{A} N)
%% \end{align*}

\input{latex/cpp2024/StepIndexedLogic.tex}
